---
title: "Test data preprocessing"
author: "Jennifer Crockett"
date: "2024-03-08"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(biomaRt) # for a version that doesn't error out with latest tidyverse update - BiocManager::install('grimbough/biomaRt')
```

# Purpose

Preprocessing the test dataset, to convert mouse gene names to human gene equivalents. GSEA only accepts human gene names, and by way of evolutionary conservation, most mouse genes have a human counterpart with a highly similar biological function.

# Load data

Load the raw gene expression dataset, downloaded from https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE72737

See `GSEAforR/testing/Analysis_using_GSEA_for_R.Rmd` for more information about this dataset.

```{r}
raw_expr <- read_csv(here("testing", "data", "GSE72737_Dnmt3a_KI_counts_raw.csv.gz"), show_col_types = F) %>%
  dplyr::rename(ensembl_gene_id=1)
```

Load the sample metadata, describing the experimental group each sample belongs to.

```{r}
meta_raw <- read_csv(here("testing", "data", "GSE72737_raw_metadata.csv"), show_col_types = F)
```


# Gene name transformation

Convert mouse gene identifiers to human gene names using the Biomart database.

```{r}
human <- useMart("ensembl", dataset = "hsapiens_gene_ensembl", host = "https://dec2021.archive.ensembl.org/")
mouse <- useMart("ensembl", dataset = "mmusculus_gene_ensembl", host = "https://dec2021.archive.ensembl.org/")

m2h <- getLDS(attributes = "ensembl_gene_id", filters = "ensembl_gene_id", values = raw_expr$ensembl_gene_id, mart = mouse,
              attributesL = "hgnc_symbol", martL = human, uniqueRows = T)

m2h <- dplyr::rename(m2h, ensembl_gene_id = Gene.stable.ID, hgnc_symbol = HGNC.symbol)
```

```{r}
# how many genes have human equivalents?
(n_equivalent <- sum(!is.na(m2h$hgnc_symbol)))

# fraction of input mouse genes with human equivalents?
n_equivalent/nrow(raw_expr)

# the rest are likely non-equivalent (non-conserved) genes
```


```{r}
expr <- left_join(raw_expr, m2h, by = "ensembl_gene_id") %>%
  filter(!is.na(hgnc_symbol)) %>%
  dplyr::select(hgnc_symbol, everything(), -ensembl_gene_id)
```

# Sample name processing

To improve the readability, refer to the metadata and update the sample names. We will name them based on their experimental group membership.

```{r}
# create unique sample names based on experimental groups
meta_conversion <- mutate(meta_raw, sample_id = make.unique(group, sep = "_")) %>%
  dplyr::select(-geo_id)

# convert sample names in the expression data
# check that the samples are in the same order for the expression data and metadata before converting
if (identical(colnames(expr[,2:ncol(expr)]), meta_conversion$description, attrib.as.set = F)) {
  new_colnames <- c(colnames(expr[,1]), meta_conversion$sample_id)
  colnames(expr) <- new_colnames
} else {
  stop("Samples are not in the same order. Fix and then try sample name conversion again.")
}
```


# Save processed data

```{r}
# expression data
write_csv(expr, here("testing", "data", "GSE72737_preprocessed_counts_human_HGNC_symbols.csv"))

# metadata
meta <- dplyr::select(meta_conversion, sample_id, group)
write_csv(meta, here("testing", "data", "GSE72737_preprocessed_metadata.csv"))
```

